Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood

### 2025-10-14 paper/paper.Rmd
add detail on vanilla trials, fix figure formatting in main and SI references

```{r growth, fig.cap="Growth curves from a logistic growth model showing predicted vocabulary growth for children with initial reaction times one SD faster than the mean (blue), at the mean (red), and one SD slower than the mean (green). Individual longitudinal trajectories are shown in light gray. Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4, fig.width=6, out.width=".6\\textwidth", include=TRUE}

@@

We included information from `r length(unique(d_sub$subject_id))` unique participants across `r length(unique(d_sub$dataset_name))` datasets. Dataset information is given in Table \ref{tab:datasets}. Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed only data from simple word recognition trials; these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations. We focus here on English purely for practical reasons -- the Peekbank dataset at present contains limited data from other languages.

@@

We used `lme4` to fit linear mixed-effects models, `brms` to fit non-linear growth models, and `lavaan` to fit structural equation models. Random effects structures for each model are given in text; full model specifications are available in the Supplemental Information and in the reproducible code for this paper, available in the linked repository. To aid interpretability, all variables were standardized (z-scored) prior to inclusion in structural equation models.

>>>>

```{r growth, fig.cap="Growth curves from a logistic growth model showing predicted vocabulary growth for children with initial reaction times one SD faster than the mean (blue), at the mean (red), and one SD slower than the mean (green). Individual longitudinal trajectories are shown in light gray. Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4, include=TRUE}

@@

We included information from `r length(unique(d_sub$subject_id))` unique participants across `r length(unique(d_sub$dataset_name))` datasets. Dataset information is given in Table \ref{tab:datasets}. Although experiments in Peekbank include a variety of different experimental manipulations, we analyzed only data from standard, simple word recognition trials; these trials were sometimes the main focus of the original studies and sometimes constituted control conditions for experiments with more complex manipulations. Requirements for being considered a standard word recognition trial included that (a) the target word was familiar (also no part-words); (b) the target word was the first point of disambiguation; (c) the target word was embedded in a well-formed, grammatical carrier phrase; (d) there was no informative language presented prenominally (e.g. semantically informative verbs, adjectives); (e) there were no nonsense words presented anywhere during the trial (including the carrier phrase); (f) there was no language-, speaker-, or accent-switching within trial; (g) the distractor image was unrelated to both the target label and the target image; and (h) there was no phonological overlap between the distractor label and target label. We focus here on English purely for practical reasons -- the Peekbank dataset at present contains limited data from other languages.

@@

We used `lme4` to fit linear mixed-effects models, `brms` to fit non-linear growth models, and `lavaan` to fit structural equation models. Random effects structures for each model are given in text; full model specifications are available in the Supplemental Information (S8, S9, and S12) and in the reproducible code for this paper, available in the linked repository. To aid interpretability, all variables were standardized (z-scored) prior to inclusion in structural equation models.

### 2025-10-14 paper/supplemental_information.Rmd
add detail on vanilla trials, fix figure formatting in main and SI references

```{r nlme-residual, fig.cap="\\label{fig:nlme-resid} Growth curves from a logistic growth model showing predicted vocabulary growth for children based on their age-residualized initial reaction times. Predictions are shown for children with initial reaction times one SD faster than the mean for their age (blue), at the mean  for their age (red), and one SD slower than the mean for their age (green). Individual longitudinal trajectories are shown in light gray. Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4, fig.width=6, out.width=".6\\textwidth", fig.pos="h!"}

@@

Our data are MAR (missing at random) rather than MCAR (missing completely at random). This is because their missingness is due to which dataset they are part of – if they are from a cross-sectional dataset, they are by definition missing all longitudinal observations. For our analyses to be appropriate given this structure, we have to assume that the general developmental patterns we are studying are replicated across datasets. We believe that they are, and we show this statistically using our mixed-effects and non-linear mixed-effects models, which control for dataset-related variation. (We also show dataset-level effects in a number of our visualizations for this same reason). The same degree of random effect specification that we can do in the mixed-effects models is not possible in the SEM model, however, purely for technical reasons. Again, this point highlights the importance of convergence across analyses.

@@

Baayen, R. H., & Milin, P. (2010). Analyzing reaction times. International journal of psychological research, 3(2), 12-28.

Creel, S. (2024). Connecting the tots: Strong looking-pointing correlations in preschoolers' word learning and implications for continuity in language development. Child Development, 96, 87-103.

Logan, G. D. (1988). Toward an instance theory of automatization. Psychological review, 95(4), 492.

Luce, P. A. (1986). A computational analysis of uniqueness points in auditory word recognition. Perception & Psychophysics, 39(3), 155-158.

Ratcliff, R. (1979). Group reaction time distributions and an analysis of distribution statistics. Psychological bulletin, 86(3), 446.

Ratcliff, R. (1993). Methods for dealing with reaction time outliers. Psychological bulletin, 114(3), 510.

Van Zandt, T. (2000). How to fit a response time distribution. Psychonomic bulletin & review, 7(3), 424-465.

Van Zandt, T. (2002). Analysis of response time distributions. Stevens’ handbook of experimental psychology, 4, 461-516.

>>>>

```{r nlme-residual, fig.cap="\\label{fig:nlme-resid} Growth curves from a logistic growth model showing predicted vocabulary growth for children based on their age-residualized initial reaction times. Predictions are shown for children with initial reaction times one SD faster than the mean for their age (blue), at the mean  for their age (red), and one SD slower than the mean for their age (green). Individual longitudinal trajectories are shown in light gray. Solid lines show global model estimates and colored regions indicate 95% credible intervals.", fig.env="figure", fig.height = 4}

@@

Our data are MAR (missing at random) rather than MCAR (missing completely at random). This is because their missingness is due to which dataset they are part of – if they are from a cross-sectional dataset, they are by definition missing all longitudinal observations. For our analyses to be appropriate given this structure, we have to assume that the general developmental patterns we are studying are replicated across datasets. We believe that they are, and we show this statistically using our mixed-effects and non-linear mixed-effects models, which control for dataset-related variation. We also show dataset-level effects in a number of our visualizations for this same reason. The same degree of random effect specification that we can do in the mixed-effects models is not possible in the SEM model, however, purely for technical reasons. Again, this point highlights the importance of convergence across analyses.

@@

Baayen, R. H., & Milin, P. (2010). Analyzing reaction times. *International Journal of Psychological Research, 3*(2), 12-28.

Creel, S. (2024). Connecting the tots: Strong looking-pointing correlations in preschoolers' word learning and implications for continuity in language development. *Child Development, 96*, 87-103.

Logan, G. D. (1988). Toward an instance theory of automatization. *Psychological Review, 95*(4), 492–527.

Luce, P. A. (1986). A computational analysis of uniqueness points in auditory word recognition. *Perception & Psychophysics, 39*(3), 155-158.

Ratcliff, R. (1979). Group reaction time distributions and an analysis of distribution statistics. *Psychological bulletin, 86*(3), 446–461.

Ratcliff, R. (1993). Methods for dealing with reaction time outliers. *Psychological Bulletin, 114*(3), 510-532.

Van Zandt, T. (2000). How to fit a response time distribution. *Psychonomic Bulletin & Review, 7*(3), 424-465.

Van Zandt, T. (2002). Analysis of response time distributions. *Stevens’ Handbook of Experimental Psychology, 4*, 461-516.
